---
title: "Initial Exploration"
format: html
editor: visual
---

```{r}
#| message: false
library(here)
library(tidyverse)
library(segmented)
```

Questions

(i) does the amount of [chemical name] increase the amount of [chemical name] contamination in macroinvertebrates?
(ii) is the risk of [chemical name] contamination/ingestion constant across the whole range of concentrations in sediment and water?
(iii) if the risk is sediment/water-concentration dependent, does a threshold value exist?


What I want to do:
- find effect independent of species



```{r}
# Loading in data
biota <- read_csv(here("data/biota.csv"))
sediment <- read_csv(here("data/sediment.csv"))
waterquality <- read_csv(here("data/waterquality.csv"))
```

```{r}
# Joining biota and sediment data
plot1 <- inner_join(biota, sediment, by = "Station_ID",
                    relationship = "many-to-many")
```


Chromium
```{r}
# Exploratory plot
ggplot(data = plot1, aes(x = Chromium.y, y = Chromium.x)) +
  geom_point() +
  labs(title = "Relationship between sediment chromium and chromium in invertebrates",
       subtitle = "Upper Clark Fork Basin, Montana",
       x = "Chromium concentration of sediment",
       y = "Chromium concentration of biota")
```

```{r}
# Picking a random taxa to explore
unique(plot1$Taxa)

hydro_occ <- plot1 %>% 
  filter(Taxa == "Hydropsyche occidentalis")
```

```{r}
# Hydropsyche occidentalis plot
ggplot(data = hydro_occ, aes(x = Chromium.y, y = Chromium.x)) +
  geom_point() +
  labs(title = "Chromium Conc in the Western Net-spinning Caddisfly",
       subtitle = "Upper Clark Fork Basin, Montana",
       x = "Chromium concentration of sediment",
       y = "Chromium concentration of biota") +
  geom_smooth(method = "lm")
```

Nickel
```{r}
# Exploratory plot
ggplot(data = plot1, aes(x = Nickel.y, y = Nickel.x)) +
  geom_point() +
  labs(title = "Relationship between sediment nickel and nickel in invertebrates",
       subtitle = "Upper Clark Fork Basin, Montana",
       x = "Nickel concentration of sediment",
       y = "Nickel concentration of biota")
```

```{r}
# Selecting one species
acrto <- plot1 %>% 
  filter(Taxa == "Arctopsyche grandis")

ggplot(data = acrto, aes(x = Chromium.y, y = Chromium.x)) +
  geom_point() +
  labs(title = "Nickel Conc in the Great Net-spinning Caddisfly",
       subtitle = "Upper Clark Fork Basin, Montana",
       x = "Nickel concentration of sediment",
       y = "Nickel concentration of biota") +
  geom_smooth(method = "lm")
```




```{r}
x <- seq(from = 1, to = 100)

m <- 50 # true mean
v <- 25  # true variance
y <- rgamma(n = 100, shape = (m^2)/v, scale = v/m)

simulated_data <- data.frame(x, y)

model <- glm(y ~ 1, 
          data = simulated_data,
          family = Gamma(link = "identity"))

summary <- summary(model)

shape_model <- 1/summary$dispersion
shape_model

scale_model <- as.numeric(coef(summary))/shape_model
scale_model
```

```{r}
# shape
shape_input_water <- ((mean(conc_arsenic$concentration_water))^2)/var(conc_arsenic$concentration_water)
shape_input_water

# scale
scale_input_water <- var(conc_arsenic$concentration_water/mean(conc_arsenic$concentration_water))
scale_input_water

# shape
shape_input_sed <- ((mean(conc_arsenic$concentration_sediment))^2)/var(conc_arsenic$concentration_sediment)
shape_input_sed

# scale
scale_input_sed <- var(conc_arsenic$concentration_sediment/mean(conc_arsenic$concentration_sediment))
scale_input_sed
```

```{r}
conc_model_gamma_summary <- summary(conc_model_gamma)

conc_model_gamma$disperson
```


```{r}
y <- conc_arsenic$concentration_biota
shape_data <- mean(y)^2 / var(y)
scale_data <- var(y) / mean(y)

> shape_data
[1] 0.9844852
> scale_data
[1] 8.173274
```


```{r}

```


summary(conc_model_gamma)$dispersion
[1] 0.2678039
> shape_hat <- 1 / summary(conc_model_gamma)$dispersion
> shape_hat
[1] 3.734075

> gamma.shape(conc_model_gamma)