---
title: "Blog post"
author: Sofia Sarak
data: November 31, 2025
format: html
editor: visual
---

# Exploring Pollutant Accumulation in Invertebrate Based on Concentrations in Water and Sediment

## Using Gamma and Segmented/Breakpoint Statistical Analyses

## Water-Quality, Bed-Sediment, and Invertebrate Tissue Trace-Element Concentrations for Tributaries in the Clark Fork Basin, Montana

Explain where data is from

Describe data set - area where the data is from and industries polluting it - relationship between water concentration and tissue concentration - harmful effects of constituents


## Why Gamma and Segmented/Breakpoint Models?

The **segmented/break-point** (or sometimes even referred to as **broken-line**) statistical model is a regression analysis where the relationship between the response and one or more predictor variables contain a break-point at which the relationship between the response and predictor variables changes. This sort of model is common in fields such as epidemiology, occupational medicine, toxicology, and ecology, where it is of interest to assess threshold value where the effect of the covariate changes.

We will will use model in our analysis to explore whether their is a threshold value(s) at which the relationship between trace-element concentration in water and trace-element accumulation in invertebrates changes.

Y∼Gamma(μ, k),log(μ)=β0​+β1​x+β2​(x−τ)

The segmented model must be applied to an already existing linear regression model, as it builds off of an already-analyzed relationship between the response and predictor variable(s) -- segmented analysis simply determines where the analyzed relationship changes.

Therefore, for the first step of our analysis we will use a **gamma distribution** statistical model. It is appropriate for data that contains continuous variables that are always positive and have skewed distributions. Our data matches this description, as trace-element concentrations are never negative (and our data contains no values of 0), and higher concentrations occur less frequently than lower ones.

The gamma distribution is made up of the parameters shape and scale, which relate to the mean and variance of the data.


Through our multi-step analysis, we aim to answer the following questions:

1. Does trace-element concentration in water increase trace-element concentration in macroinvertebrates?
2. Is the risk of trace-element contamination/ingestion constant across the whole range of concentrations in sediment and water?
3. If the risk is sediment/water-concentration dependent, does a threshold value exist?


**Null Hypothesis:** There is no relationship between trace-element concentration in water and biota, and therefore a breakpoint doesn't exist.

**Alternative Hypthesis:** There is a positive relationship between trace-element concentration in water and biota, and a breakpoint does exist.

## The Analysis

```{r}
#| message: false
library(here)
library(tidyverse)
library(janitor)
library(ggplot2)
library(segmented)
```

```{r}
#| message: false
# Loading in data
biota <- read_csv(here("data/biota.csv")) %>% 
  clean_names()
sediment <- read_csv(here("data/sediment.csv")) %>% 
  clean_names()
waterquality <- read_csv(here("data/waterquality.csv")) %>% 
  clean_names()
```

```{r}
#| warning: false
waterquality <- waterquality %>% 
  mutate(date = mdy(date), 
         month = month(date, label = TRUE),
         year = year(date)) %>% 
  relocate(c(month, year), .after = date) %>% 
  filter(month == "Jul") %>% 
  filter(year == 2021) %>% 
  dplyr::select(-contains("rc_")) %>% 
  pivot_longer(cols = nitrate_plus_nitrite:suspended_sediment_concentration,
               names_to = "constituent",
               values_to = "concentration_water") %>% 
   mutate(station_id = as.character(station_id))

waterquality$constituent <- gsub("_wf$", "", waterquality$constituent)
waterquality$constituent <- gsub("_wu$", "", waterquality$constituent)

sediment <- sediment %>% 
  dplyr::select(-contains(c("rc_", "flag"))) %>% 
  pivot_longer(cols = arsenic:zinc,
               names_to = "constituent",
               values_to = "concentration_sediment") %>% 
  mutate(station_id = as.character(station_id))

biota <- biota %>% 
  dplyr::select(-contains(c("rc_", "flag"))) %>% 
  mutate(lead = as.numeric(lead)) %>% 
  pivot_longer(cols = arsenic:zinc,
               names_to = "constituent",
               values_to = "concentration_biota") %>% 
  mutate(station_id = as.character(station_id))
```

```{r}
waterquality_filtered <- waterquality %>%
  filter(constituent %in% sediment$constituent)

enviro <- inner_join(waterquality_filtered, sediment, by = c("constituent", "station_id"), 
                     relationship = "many-to-many")
conc <- inner_join(enviro, biota, by = c("constituent", "station_id"),
                   relationship = "many-to-many")
```

Models for arsenic
```{r}
conc_arsenic <- conc %>% 
  filter(constituent == "arsenic")

conc_arsenic <- conc_arsenic %>% 
  mutate(sediment_conc_bins = cut(conc_arsenic$concentration_sediment,breaks = c(0, 50, 100, 150)))

conc_model_gamma <- glm(concentration_biota ~ concentration_water + concentration_sediment, 
                   data = conc_arsenic, # data remains unchanged
                   family = Gamma(link = "log")) # family and link function

conc_model_seg <- segmented(obj = conc_model_gamma, seg.Z = ~concentration_water + concentration_sediment)
```

Creating simulated data
```{r}
# Create prediction grid with levels of water and sediment concentrations
pred_grid <- expand_grid(concentration_water = 1:30,
                         concentration_sediment = 1:150)
```

Applying gamma model to simulated values
```{r}
conc_model_gamma_se <- predict(object = conc_model_gamma,
                          newdata = pred_grid,
                          type = "link", # stay in link space
                          se.fit = TRUE)


concentration_pred <- pred_grid %>% 
  mutate(log_conc = conc_model_se$fit,
    # 95% CI in link space
    log_conc_se = conc_model_se$se.fit,
    log_conc_lwr = qnorm(0.025, mean = log_conc, sd = log_conc_se),
    log_conc_upr = qnorm(0.975, mean = log_conc, sd = log_conc_se),
    # undo the link function -- in this case log --  using exp()
    conc = exp(log_conc),
    conc_lwr = exp(log_conc_lwr),
    conc_upr = exp(log_conc_upr))
```

Gamma Plot
```{r}
conc_model_pred_filtered <- concentration_pred %>% 
  filter(concentration_sediment %in% c(50, 100, 150))

ggplot(conc_model_pred_filtered, aes(x = concentration_water, y = conc)) +
  geom_line(aes(color = as_factor(concentration_sediment))) +
  geom_ribbon(aes(ymin = conc_lwr, ymax = conc_upr, fill = as_factor(concentration_sediment)),
              alpha = 0.2) + 
  geom_point(data = conc_arsenic, aes(x = concentration_water, y = concentration_biota, color = sediment_conc_bins)) +
  labs(title = "Gamma distribution") +
  xlim(0, 25) +
  ylim(0, 100) 
```

Applying segmented model to simulated values
```{r}
conc_model_seg_se <- predict(object = conc_model_seg,
                          newdata = pred_grid,
                          type = "link", # stay in link space
                          se.fit = TRUE)

concentration_pred_seg <- pred_grid %>% 
  mutate(log_conc_seg = conc_model_seg_se$fit,
    # 95% CI in link space
    log_conc_seg_se = conc_model_seg_se$se.fit,
    log_conc_seg_lwr = qnorm(0.025, mean = log_conc_seg, sd = log_conc_seg_se),
    log_conc_seg_upr = qnorm(0.975, mean = log_conc_seg, sd = log_conc_seg_se),
    # undo the link function -- in this case log --  using exp()
    conc = exp(log_conc_seg),
    conc_lwr = exp(log_conc_seg_lwr),
    conc_upr = exp(log_conc_seg_upr))
```

Creating segmented plot
```{r}
conc_pred_seg_filtered <- concentration_pred_seg %>% 
  filter(concentration_sediment %in% c(50, 100, 150))

ggplot(data = conc_pred_seg_filtered, aes(x = concentration_water, y = conc)) +
  geom_line(aes(color = as_factor(concentration_sediment))) +
  geom_point(data = conc_arsenic, aes(x = concentration_water, y = concentration_biota, color = sediment_conc_bins)) +
  geom_ribbon(data = conc_pred_seg_filtered, aes(ymin = conc_lwr, ymax = conc_upr, fill = as_factor(concentration_sediment)),
              alpha = 0.2) +
  xlim(0, 25) +
  ylim(0, 100) 
```

## Interpretation

Visually, it appears that both a positive relationship and a breakpoint exist between trace-element concentration in water and biota. To confirm, we can take a look at the model summary ouputs.


