---
title: "Blog post"
format: html
editor: visual
---

```{r}
#| message: false
library(here)
library(tidyverse)
library(janitor)
library(ggplot2)
library(segmented)
```

```{r}
#| message: false
# Loading in data
biota <- read_csv(here("data/biota.csv")) %>% 
  clean_names()
sediment <- read_csv(here("data/sediment.csv")) %>% 
  clean_names()
waterquality <- read_csv(here("data/waterquality.csv")) %>% 
  clean_names()
```

Describe data set
- area where the data is from and industries polluting it
- relationship between water concentration and tissue concentration
- harmful effects of constituents

- what is the relationship between water concentration and bed sediment concentrations?
  - combine data sets and pivot longer:
  - rows that contain "water" or "bed-sediment", as well as concentration of constituent, and constituent as factor
  - create scatterplots with lines for water and bed-sediment, faceted by constituent
  
```{r}
waterquality <- waterquality %>% 
  mutate(date = mdy(date), 
         month = month(date, label = TRUE),
         year = year(date)) %>% 
  relocate(c(month, year), .after = date) %>% 
  filter(month == "Jul") %>% 
  filter(year == 2021) %>% 
  dplyr::select(-contains("rc_")) %>% 
  pivot_longer(cols = nitrate_plus_nitrite:suspended_sediment_concentration,
               names_to = "constituent",
               values_to = "concentration_water") %>% 
   mutate(station_id = as.character(station_id))

waterquality$constituent <- gsub("_wf$", "", waterquality$constituent)
waterquality$constituent <- gsub("_wu$", "", waterquality$constituent)

sediment <- sediment %>% 
  dplyr::select(-contains(c("rc_", "flag"))) %>% 
  pivot_longer(cols = arsenic:zinc,
               names_to = "constituent",
               values_to = "concentration_sediment") %>% 
  mutate(station_id = as.character(station_id))

biota <- biota %>% 
  dplyr::select(-contains(c("rc_", "flag"))) %>% 
  mutate(lead = as.numeric(lead)) %>% 
  pivot_longer(cols = arsenic:zinc,
               names_to = "constituent",
               values_to = "concentration_biota") %>% 
  mutate(station_id = as.character(station_id))
```

```{r}
waterquality_filtered <- waterquality %>%
  filter(constituent %in% sediment$constituent)

enviro <- inner_join(waterquality_filtered, sediment, by = c("constituent", "station_id"), 
                     relationship = "many-to-many")
conc <- inner_join(enviro, biota, by = c("constituent", "station_id"),
                   relationship = "many-to-many")
```

Describe statistical models
- gamma
    - shape and scale, and how they relate to mean and variance
- segmented


Create model
```{r}
conc_arsenic <- conc %>% 
  filter(constituent == "arsenic")

conc_model_gamma <- glm(concentration_biota ~ concentration_water + concentration_sediment, 
                   data = conc_arsenic, # data remains unchanged
                   family = Gamma(link = "log")) # family and link function

conc_model_seg <- segmented(obj = conc_model_gamma, seg.Z = ~concentration_water, psi = 10)
```


Fit simulated data to models
- simulate data using rgamma (and mean and variance, transformed into shape and scale)

```{r}
# Create prediction grid with levels of water and sediment concentrations
pred_grid <- expand_grid(concentration_water = 1:30,
                         concentration_sediment = 1:150)

concentration_pred <- pred_grid %>% 
  mutate(biota_conc_pred_guass = predict(object = concentration_model,
                                        newdata = pred_grid,
                                        type = "link"),
         biota_conc_pred_seg = predict(object = conc_model_seg,
                                       newdata = pred_grid,
                                       type = "link") )
```

```{r}
conc_model_se <- predict(object = conc_model_gauss,
                          newdata = pred_grid,
                          type = "link", # stay in link space
                          se.fit = TRUE)


concentration_pred <- pred_grid %>% 
  mutate(log_conc = conc_model_se$fit,
    # 95% CI in link space
    log_conc_se = conc_model_se$se.fit,
    log_conc_lwr = qnorm(0.025, mean = log_conc, sd = log_conc_se),
    log_conc_upr = qnorm(0.975, mean = log_conc, sd = log_conc_se),
    # undo the link function using linkinv
    conc = exp(log_conc),
    conc_lwr = exp(log_conc_lwr),
    conc_upr = exp(log_conc_upr))
```

```{r}
ggplot(concentration_pred, aes(x = concentration_water, y = conc)) +
  geom_ribbon(aes(ymin = conc_lwr, ymax = conc_upr), alpha = 0.2) +
  geom_line()
```

```{r}
conc_model_pred_filtered <- concentration_pred %>% 
  filter(concentration_sediment %in% c(50, 100, 150))

ggplot(conc_model_pred_filtered, aes(x = concentration_water, y = conc)) +
  geom_line(aes(color = as_factor(concentration_sediment))) +
  geom_ribbon(aes(ymin = conc_lwr, ymax = conc_upr, fill = as_factor(concentration_sediment)),
              alpha = 0.2)
```


To do
- figure out how to plot segmented model
- and plot with original data


- using what I did below, confirm that model returns mean and variance of data





```{r}
x <- seq(from = 1, to = 100)

m <- 50 # true mean
v <- 25  # true variance
y <- rgamma(n = 100, shape = (m^2)/v, scale = v/m)

simulated_data <- data.frame(x, y)

model <- glm(y ~ 1, 
          data = simulated_data,
          family = Gamma(link = "identity"))

summary <- summary(model)

shape_model <- 1/summary$dispersion
shape_model

scale_model <- as.numeric(coef(summary))/shape_model
scale_model
```

```{r}
# shape
shape_input <- (mean(simulated_data$y)^2)/var(simulated_data$y)
shape_input

# scale
scale_input <- var(simulated_data$y/mean(simulated_data$y))
scale_input
```



```{r}
ggplot(simulated_data, aes(y)) +
  geom_density()
```
```{r}

```


```{r}
 segmented.glm(obj = fit.glm, seg.Z = ~age, psi = 25)
```


- plot data from rgamma
- use simulated data and run a segmented model
- plot line on top of scatter plot

Fit real data to models
- use water concentrations, controlling for sediment concentration
  - using the assumption that water is the main thing affecting tissue constituent concentration
- calculate for arsenic
- create function that takes in name of constituent