---
title: "Blog post"
format: html
editor: visual
---

```{r}
#| message: false
library(here)
library(tidyverse)
library(janitor)
library(ggplot2)
library(segmented)
```

```{r}
#| message: false
# Loading in data
biota <- read_csv(here("data/biota.csv")) %>% 
  clean_names()
sediment <- read_csv(here("data/sediment.csv")) %>% 
  clean_names()
waterquality <- read_csv(here("data/waterquality.csv")) %>% 
  clean_names()
```

Describe data set
- area where the data is from and industries polluting it
- relationship between water concentration and tissue concentration
- harmful effects of constituents

- what is the relationship between water concentration and bed sediment concentrations?
  - combine data sets and pivot longer:
  - rows that contain "water" or "bed-sediment", as well as concentration of constituent, and constituent as factor
  - create scatterplots with lines for water and bed-sediment, faceted by constituent
  
```{r}
waterquality <- waterquality %>% 
  mutate(date = mdy(date), 
         month = month(date, label = TRUE),
         year = year(date)) %>% 
  relocate(c(month, year), .after = date) %>% 
  filter(month == "Jul") %>% 
  filter(year == 2021) %>% 
  dplyr::select(-contains("rc_")) %>% 
  pivot_longer(cols = nitrate_plus_nitrite:suspended_sediment_concentration,
               names_to = "constituent",
               values_to = "concentration_water")

sediment <- sediment %>% 
  dplyr::select(-contains(c("rc_", "flag"))) %>% 
  pivot_longer(cols = arsenic:zinc,
               names_to = "constituent",
               values_to = "concentration_sediment")

biota <- biota %>% 
  dplyr::select(-contains(c("rc_", "flag"))) %>% 
  mutate(lead = as.numeric(lead)) %>% 
  pivot_longer(cols = arsenic:zinc,
               names_to = "constituent",
               values_to = "concentration_biota")
```


```{r}
enviro <- inner_join(waterquality, sediment, by = "site_name",
                     relationship = "many-to-many")
```

to do:
- in water quality, remove _wf from constituent factors

Describe statistical models
- gamma
    - shape and scale, and how they relate to mean and variance
- segmented

Fit simulated data to models
- simulate data using rgamma (and mean and variance, transformed into shape and scale)

```{r}
x <- seq(from = 1, to = 100)

m <- 50 # true mean
v <- 25  # true variance
y <- rgamma(n = 100, shape = (m^2)/v, scale = v/m)

simulated_data <- data.frame(x, y)

ggplot(simulated_data, aes(y)) +
  geom_histogram()
```
```{r}
model <- glm(y ~ x, 
          data = simulated_data,
          family = Gamma(link = "identity"))
```


```{r}
 segmented.glm(obj = fit.glm, seg.Z = ~age, psi = 25)
```


- plot data from rgamma
- use simulated data and run a segmented model
- plot line on top of scatter plot

Fit real data to models
- use water concentrations, controlling for sediment concentration
  - using the assumption that water is the main thing affecting tissue constituent concentration
- calculate for arsenic
- create function that takes in name of constituent