---
title: "Blog post"
author: Sofia Sarak
data: November 31, 2025
format: html
editor: visual
---

# Exploring Pollutant Accumulation in Invertebrate Based on Concentrations in Water and Sediment

## Using Gamma and Segmented/Breakpoint Statistical Analyses

## Water-Quality, Bed-Sediment, and Invertebrate Tissue Trace-Element Concentrations for Tributaries in the Clark Fork Basin, Montana

Explain where data is from

Describe data set - area where the data is from and industries polluting it - relationship between water concentration and tissue concentration - harmful effects of constituents


## Why Gamma and Segmented/Breakpoint Models?

### Model 1: Segmented

The **segmented/break-point** (or sometimes even referred to as **broken-line**) statistical model is a regression analysis where the relationship between the response and one or more predictor variables contain a break-point at which the relationship between the response and predictor variables changes. This sort of model is common in fields such as epidemiology, occupational medicine, toxicology, and ecology, where it is of interest to assess threshold value where the effect of the covariate changes.

We will will use model in our analysis to explore whether their is a threshold value(s) at which the relationship between trace-element concentration in water and trace-element accumulation in invertebrates changes.

To test this analysis on simulated data, we can do so by feeding it data that has a known break-point, such as that below.

```{r}
#| message: false
#| echo: false
library(here)
library(tidyverse)
library(janitor)
library(ggplot2)
library(segmented)
library(patchwork)
```

```{r}
#| warning: false
#| echo: false
# Simulate data frame with known break-point of 50
sample_data <- data.frame(x = seq(1:100))

sample_data <- sample_data %>% 
  mutate(y = if_else(x < 50, true = x, false = x*2-50))

ggplot(data = sample_data, aes(x = x, y = y)) +
  geom_line(color = "red") +
  geom_vline(xintercept = 50, color = "gray", linetype = "dashed") + 
  theme_light() +
  labs(title = "Simulated Data with Known Breakpoint at x = 50")
```

To follow the steps of a segmented model, we first apply a linear regression model to the data, then the segmented model.
```{r}
# Run lm model on it
sample_model_seg_1 <- lm(data = sample_data, y~x)

# Run segmented model on lm (note: function comes from the R package "segmented")
sample_model_seg_2 <- segmented(sample_model_seg_1, seg.Z = ~x)
```

If we were successful, the summary should return an estimated breakpoint at x = 50.
```{r}
# Does it return a break-point of 50?
summary(sample_model_seg_2)
```

It does!

### Model 2: Gamma

Y∼Gamma(μ, k),log(μ)=β0​+β1​x+β2​(x−τ)

The segmented model must be applied to an already existing linear regression model, as it builds off of an already-analyzed relationship between the response and predictor variable(s) -- segmented analysis simply determines where the analyzed relationship changes.

Therefore, for the first step of our analysis we will use a **gamma distribution** statistical model. It is appropriate for data that contains continuous variables that are always positive and have skewed distributions. Our data matches this description, as trace-element concentrations are never negative (and our data contains no values of 0), and higher concentrations occur less frequently than lower ones.

The gamma distribution is made up of the parameters shape and scale, which relate to the mean and variance of the data. We test the model using simulated data below.


Through our multi-step analysis, we aim to answer the following questions:

1. Does trace-element concentration in water increase trace-element concentration in macroinvertebrates?
2. Is the risk of trace-element contamination/ingestion constant across the whole range of concentrations in sediment and water?
3. If the risk is sediment/water-concentration dependent, does a threshold value exist?


**Null Hypothesis:** There is no relationship between trace-element concentration in water and biota, and therefore a breakpoint doesn't exist.

**Alternative Hypthesis:** There is a positive relationship between trace-element concentration in water and biota, and a breakpoint does exist.

## The Analysis

```{r}
#| message: false
#| echo: false
# Loading in data
biota <- read_csv(here("data/biota.csv")) %>% 
  clean_names()
sediment <- read_csv(here("data/sediment.csv")) %>% 
  clean_names()
waterquality <- read_csv(here("data/waterquality.csv")) %>% 
  clean_names()
```

```{r}
#| warning: false
#| echo: false
waterquality <- waterquality %>% 
  mutate(date = mdy(date), 
         month = month(date, label = TRUE),
         year = year(date)) %>% 
  relocate(c(month, year), .after = date) %>% 
  filter(month == "Jul") %>% 
  filter(year == 2021) %>% 
  dplyr::select(-contains("rc_")) %>% 
  pivot_longer(cols = nitrate_plus_nitrite:suspended_sediment_concentration,
               names_to = "constituent",
               values_to = "concentration_water") %>% 
   mutate(station_id = as.character(station_id))

waterquality$constituent <- gsub("_wf$", "", waterquality$constituent)
waterquality$constituent <- gsub("_wu$", "", waterquality$constituent)

sediment <- sediment %>% 
  dplyr::select(-contains(c("rc_", "flag"))) %>% 
  pivot_longer(cols = arsenic:zinc,
               names_to = "constituent",
               values_to = "concentration_sediment") %>% 
  mutate(station_id = as.character(station_id))

biota <- biota %>% 
  dplyr::select(-contains(c("rc_", "flag"))) %>% 
  mutate(lead = as.numeric(lead)) %>% 
  pivot_longer(cols = arsenic:zinc,
               names_to = "constituent",
               values_to = "concentration_biota") %>% 
  mutate(station_id = as.character(station_id))
```

```{r}
waterquality_filtered <- waterquality %>%
  filter(constituent %in% sediment$constituent)

enviro <- inner_join(waterquality_filtered, sediment, by = c("constituent", "station_id"), 
                     relationship = "many-to-many")
conc <- inner_join(enviro, biota, by = c("constituent", "station_id"),
                   relationship = "many-to-many")
```

```{r}
water_plot <- ggplot(data = conc, aes(x = constituent, y = concentration_water)) +
  geom_boxplot(aes(fill = constituent)) +
  theme_minimal() +
  labs(x = "Trace-Element",
       y = "Concentration (µg/L)",
       title = "Water Quality") +
  scale_fill_brewer(name = "Trace-Element", palette = "Dark2") +
  theme(axis.text.x = element_text(angle = -45))
                                   
sediment_plot <- ggplot(data = conc, aes(x = constituent, y = concentration_sediment)) +
  geom_boxplot(aes(fill = constituent)) +
  theme_minimal() +
  labs(x = "Trace-Element",
       y = "Concentration (µg/L)",
       title = "Sediment") +
  scale_fill_brewer(name = "Trace-Element", palette = "Dark2") 

invert_plot <- ggplot(data = conc, aes(x = constituent, y = concentration_biota)) +
  geom_boxplot(aes(fill = constituent)) +
  theme_minimal() +
  labs(x = "Trace-Element",
       y = "Concentration (µg/L)",
       title = "Invertebrate") +
  scale_fill_brewer(name = "Trace-Element", palette = "Dark2")

all_plots <- water_plot / sediment_plot / invert_plot +
  plot_layout(axis_titles = 'collect',
              guides = 'collect')

ggsave('figures/all_plots.png', all_plots, width = 10, height = 7)

conc_plot2 <- conc %>% 
  filter(constituent %in% c("arsenic", "cadnium","copper","lead","zinc"))
```


```{r}
water_plot_2 <- ggplot(data = conc_plot2, aes(x = constituent, y = concentration_water)) +
  geom_boxplot(aes(fill = constituent)) +
  theme_minimal() +
  labs(x = "Trace-Element",
       y = "Concentration (µg/L)",
       title = "Water Quality") +
  scale_fill_brewer(name = "Trace-Element", palette = "Dark2") +
  theme(axis.text.x = element_text(angle = -45))
                                   
sediment_plot_2 <- ggplot(data = conc_plot2, aes(x = constituent, y = concentration_sediment)) +
  geom_boxplot(aes(fill = constituent)) +
  theme_minimal() +
  labs(x = "Trace-Element",
       y = "Concentration (µg/L)",
       title = "Sediment") +
  scale_fill_brewer(name = "Trace-Element", palette = "Dark2") 

invert_plot_2 <- ggplot(data = conc_plot2, aes(x = constituent, y = concentration_biota)) +
  geom_boxplot(aes(fill = constituent)) +
  theme_minimal() +
  labs(x = "Trace-Element",
       y = "Concentration (µg/L)",
       title = "Invertebrate") +
  scale_fill_brewer(name = "Trace-Element", palette = "Dark2")

all_plots_2 <- water_plot_2 / sediment_plot_2 / invert_plot_2 +
  plot_layout(axis_titles = 'collect',
              guides = 'collect')

ggsave('figures/all_plots_2.png', all_plots_2, width = 10, height = 7)
```

```{r}
ggplot(data = conc, aes(x = concentration_water, y = concentration_biota)) + 
  geom_point(size = 0.75) +
  facet_wrap(~constituent, ncol = 4, scales = "free") +
  theme_minimal() +
  theme(strip.background = element_rect("#acd0f2")) +
  labs(x = "Water (µg/L)",
       y = "Biota (µg/L)",
       title = "Trace-Element Concentrations") +
  theme(plot.title = element_text(hjust = 0.5))
```



Models for arsenic
```{r}
conc_arsenic <- conc %>% 
  filter(constituent == "arsenic")

conc_arsenic <- conc_arsenic %>% 
  mutate(sediment_conc_bins = cut(conc_arsenic$concentration_sediment,breaks = c(0, 50, 100, 150)))

conc_model_gamma <- glm(concentration_biota ~ concentration_water + concentration_sediment, 
                   data = conc_arsenic, # data remains unchanged
                   family = Gamma(link = "log")) # family and link function

conc_model_seg <- segmented(obj = conc_model_gamma, seg.Z = ~concentration_water + concentration_sediment)
```

Creating simulated data
```{r}
# Create prediction grid with levels of water and sediment concentrations
pred_grid <- expand_grid(concentration_water = 1:30,
                         concentration_sediment = 1:150)
```

Applying gamma model to simulated values
```{r}
conc_model_gamma_se <- predict(object = conc_model_gamma,
                          newdata = pred_grid,
                          type = "link", # stay in link space
                          se.fit = TRUE)


concentration_pred <- pred_grid %>% 
  mutate(log_conc = conc_model_gamma_se$fit,
    # 95% CI in link space
    log_conc_se = conc_model_gamma_se$se.fit,
    log_conc_lwr = qnorm(0.025, mean = log_conc, sd = log_conc_se),
    log_conc_upr = qnorm(0.975, mean = log_conc, sd = log_conc_se),
    # undo the link function -- in this case log --  using exp()
    conc = exp(log_conc),
    conc_lwr = exp(log_conc_lwr),
    conc_upr = exp(log_conc_upr))
```

Gamma Plot
```{r}
conc_model_pred_filtered <- concentration_pred %>% 
  filter(concentration_sediment %in% c(50, 100, 150))

ggplot(conc_model_pred_filtered, aes(x = concentration_water, y = conc)) +
  
  # Predicted values as a line, colored by sediment concentration
  geom_line(aes(color = as_factor(concentration_sediment))) +
  
  # 95% CIs as ribbon, colored by sediment concentration
  geom_ribbon(aes(ymin = conc_lwr, ymax = conc_upr, 
                  fill = as_factor(concentration_sediment)), alpha = 0.2) + 
  
  # Raw data as points, colored by sediment concentration bins
  geom_point(data = conc_arsenic, aes(x = concentration_water, 
                                      y = concentration_biota, 
                                      color = sediment_conc_bins)) +
  
  # Aesthetics
  scale_color_manual(name = "Sediment (µg/g)", values = c("seagreen", "goldenrod", "mediumorchid4", 
                                                          "seagreen", "goldenrod", "mediumorchid4")) +
  scale_fill_manual(values = c("seagreen", "goldenrod", "mediumorchid4")) +
  labs(x = "Water (µg/L)",
       y = "Biota (µg/L)",
       title = "Predicted Arsenic Concentrations (Gamma Model)",
       color = "Sediment (µg/g) Bins") +
  guides(fill = "none") +
  xlim(c(0, 25)) +
  ylim(c(0, 100)) +
  theme_minimal()
```

Applying segmented model to simulated values
```{r}
conc_model_seg_se <- predict(object = conc_model_seg,
                          newdata = pred_grid,
                          type = "link", # stay in link space
                          se.fit = TRUE)

concentration_pred_seg <- pred_grid %>% 
  mutate(log_conc_seg = conc_model_seg_se$fit,
    # 95% CI in link space
    log_conc_seg_se = conc_model_seg_se$se.fit,
    log_conc_seg_lwr = qnorm(0.025, mean = log_conc_seg, sd = log_conc_seg_se),
    log_conc_seg_upr = qnorm(0.975, mean = log_conc_seg, sd = log_conc_seg_se),
    # undo the link function -- in this case log --  using exp()
    conc = exp(log_conc_seg),
    conc_lwr = exp(log_conc_seg_lwr),
    conc_upr = exp(log_conc_seg_upr))
```

Creating segmented plot
```{r}
conc_pred_seg_filtered <- concentration_pred_seg %>% 
  filter(concentration_sediment %in% c(50, 100, 150))

ggplot(data = conc_pred_seg_filtered, aes(x = concentration_water, y = conc)) +
  geom_line(aes(color = as_factor(concentration_sediment))) +
  geom_point(data = conc_arsenic, aes(x = concentration_water, y = concentration_biota, color = sediment_conc_bins)) +
  geom_ribbon(data = conc_pred_seg_filtered, aes(ymin = conc_lwr, ymax = conc_upr, fill = as_factor(concentration_sediment)),
              alpha = 0.2) +
  
  scale_color_manual(name = "Sediment (µg/g)", values = c("seagreen", "goldenrod", "mediumorchid4", 
                                                          "seagreen", "goldenrod", "mediumorchid4")) +
  scale_fill_manual(values = c("seagreen", "goldenrod", "mediumorchid4")) +
  labs(x = "Water (µg/L)",
       y = "Biota (µg/L)",
       title = "Predicted Arsenic Concentrations (Break-Point Model)",
       color = "Sediment (µg/g) Bins") +
  guides(fill = "none") +
  xlim(0, 25) +
  ylim(0, 100) +
  theme_minimal()
```

## Interpretation

Visually, it appears that both a positive relationship and a break-point exist between trace-element concentration in water and biota. To confirm, we can take a look at the model summary ouputs.


## Generalized Function

Our initial analysis only applied the gamma and segmented model to arsenic, but this dataset contains 6 other trace-elements to explore. To speed up the process, a function can be made. The inputs to the function `trace_element_model`, are as follows:

- `trace-element`: a string containing the trace-element, from the data, you wish to model (all lowercase)
- `bins`: a vector containing the bin values (starting at 0 and ending with the max value) of the trace-element
- `conc_water_max` & `conc_sediment_max`: the maximum values that a grid of predicted values will be made with
- `filtered_sed`: which 3 values of sediment concentrations you wish to plot (should match `bins`, but without the 0)
- `xlim` & `ylim`: when plotting, what range of x's and y's should the graph include? (should be based on trace-element concentration range)

Function
```{r}

```

We can use the function to calculate predictions and create graphs for the other 6 trace elements.